### 2024-07-31 getting-started.qmd
20240731 - duplicate instances

@@

@@

@@

@@

@@

@@

@@

@@

player_stats_weekly_kicking <- player_stats_weekly_kicking %>% 
  dplyr::left_join(
    years_of_experience_offense[,c("player_id","season","years_of_experience")],
    by = c("player_id","season")
  )
```

@@

player_stats_seasonal_kicking <- player_stats_seasonal_kicking %>% 
  dplyr::left_join(
    years_of_experience_offense[,c("player_id","season","years_of_experience")],
    by = c("player_id","season")
  )
```

>>>>

The `nfl_players` object is in `id` form.
That is, each row should be uniquely identified by `gsis_id`.
Let's rearrange the data accordingly:

```{r}
nfl_players <- nfl_players %>% 
  select(gsis_id, everything()) %>% 
  arrange(display_name)
```

Let's check for duplicate `id` instances:

```{r}
nfl_players %>% 
  group_by(gsis_id) %>% 
  filter(n() > 1)
```

Let's do some data cleanup:

```{r}
nfl_players[nfl_players == ""] <- NA

nfl_players <- nfl_players %>% 
  filter(!is.na(gsis_id))
```

@@

The `nfl_teams` object is in `id` form.
That is, each row should be uniquely identified by `team_id`.
Let's rearrange the data accordingly:

```{r}
nfl_teams <- nfl_teams %>% 
  select(team_id, everything())
```

Let's check for duplicate `id` instances:

```{r}
nfl_teams %>% 
  group_by(team_id) %>% 
  filter(n() > 1)
```

@@

The `nfl_rosters` object is in `id`-`season`-`team` form.
That is, each row should be uniquely identified by the combination of `gsis_id`, `season`, and `team`.
Let's rearrange the data accordingly:

```{r}
nfl_rosters <- nfl_rosters %>% 
  select(gsis_id, season, team, week, everything()) %>% 
  arrange(full_name, gsis_id, season, team, week)
```

Let's check for duplicate `id` instances:

```{r}
nfl_rosters %>% 
  group_by(gsis_id, season, team) %>% 
  filter(n() > 1)
```

Let's do some data cleanup:

```{r}
# Drop players with missing values for gsis_id
nfl_rosters <- nfl_rosters %>% 
  filter(!is.na(gsis_id))

# Fill in missing values for a player in their duplicate instances, and then keep only the first of the duplicate instances
nfl_rosters <- nfl_rosters %>% 
  group_by(gsis_id, season, team) %>% 
  fill(names(.), .direction = "downup") %>% 
  slice_head(n = 1) %>% 
  ungroup()
```

Let's check again for duplicate `id` instances:

```{r}
nfl_rosters %>% 
  group_by(gsis_id, season, team) %>% 
  filter(n() > 1)
```

The `nfl_rosters_weekly` object is in `id`-`season`-`week` form.
That is, each row should be uniquely identified by the combination of `gsis_id`, `season`, and `week`.
Let's rearrange the data accordingly:

```{r}
nfl_rosters_weekly <- nfl_rosters_weekly %>% 
  select(gsis_id, season, week, everything()) %>% 
  arrange(full_name, gsis_id, season, week)
```

Let's check for duplicate `id` instances:

```{r}
nfl_rosters_weekly %>% 
  group_by(gsis_id, season, week) %>% 
  filter(n() > 1)
```

Let's do some data cleanup:

```{r}
# Drop players with missing values for gsis_id
nfl_rosters_weekly <- nfl_rosters_weekly %>% 
  filter(!is.na(gsis_id))

# Fill in missing values for a player in their duplicate instances, and then keep only the first of the duplicate instances
nfl_rosters_weekly <- nfl_rosters_weekly %>% 
  group_by(gsis_id, season, week) %>% 
  fill(names(.), .direction = "downup") %>% 
  slice_head(n = 1) %>% 
  ungroup()
```

Let's check again for duplicate `id` instances:

```{r}
nfl_rosters_weekly %>% 
  group_by(gsis_id, season, week) %>% 
  filter(n() > 1)
```

@@

The `nfl_schedules` object is in `id` form.
That is, each row should be uniquely identified by `game_id`.

Let's check for duplicate `id` instances:

```{r}
nfl_schedules %>% 
  group_by(game_id) %>% 
  filter(n() > 1)
```

@@

The `nfl_combine` object is in `id` form.
That is, each row should be uniquely identified by their `id`.
However, there is no `gsis_id` variable to merge it easily with other datasets.
Some of the players have other `id` variables, including `pfr_id` and `cfb_id`.
Let's rearrange the data accordingly:

```{r}
nfl_combine <- nfl_combine %>% 
  select(pfr_id, cfb_id, everything()) %>% 
  arrange(season, player_name)
```

Let's check for duplicate `id` instances:

```{r}
nfl_combine %>% 
  group_by(pfr_id) %>% 
  filter(n() > 1, !is.na(pfr_id)) %>% 
  arrange(pfr_id)

nfl_combine %>% 
  group_by(cfb_id) %>% 
  filter(n() > 1, !is.na(cfb_id)) %>% 
  arrange(cfb_id)
```

However, these apparent duplicates appear to be different players:

```{r}
nfl_combine %>% 
  group_by(season, pfr_id, pos) %>% 
  filter(n() > 1, !is.na(pfr_id)) %>% 
  arrange(pfr_id)

nfl_combine %>% 
  group_by(season, cfb_id, pos) %>% 
  filter(n() > 1, !is.na(cfb_id)) %>% 
  arrange(cfb_id)
```

@@

The `nfl_draftPicks` object is in `id` form.
That is, each row should be uniquely identified by `gsis_id`.
Let's rearrange the data accordingly:

```{r}
nfl_draftPicks <- nfl_draftPicks %>% 
  select(gsis_id, everything()) %>% 
  arrange(pfr_player_name)
```

Let's check for duplicate `id` instances:

```{r}
nfl_draftPicks %>% 
  group_by(gsis_id) %>% 
  filter(n() > 1)
```

Let's do some data cleanup:

```{r}
nfl_draftPicks[nfl_draftPicks == ""] <- NA

nfl_draftPicks <- nfl_draftPicks %>% 
  filter(!is.na(gsis_id))
```

Let's check again for duplicate `id` instances:

```{r}
nfl_draftPicks %>% 
  group_by(gsis_id) %>% 
  filter(n() > 1)
```

@@

The `nfl_depthCharts` object is in `id`-`season`-`week`-`position` form.
That is, each row should be uniquely identified by the combination of `gsis_id`, `season`, `week`, and `depth_position`.
Let's rearrange the data accordingly:

```{r}
nfl_depthCharts <- nfl_depthCharts %>% 
  select(gsis_id, season, week, depth_position, everything()) %>% 
  arrange(full_name, gsis_id, season, week, depth_position)
```

Let's check for duplicate `id` instances:

```{r}
nfl_depthCharts %>% 
  group_by(gsis_id, season, week, depth_position) %>% 
  filter(n() > 1)
```

Let's do some data cleanup:

```{r}
# Drop players with missing values for gsis_id
nfl_depthCharts <- nfl_depthCharts %>% 
  filter(!is.na(gsis_id))

# Fill in missing values for a player in their duplicate instances, and then keep only the first of the duplicate instances
nfl_depthCharts <- nfl_depthCharts %>% 
  group_by(gsis_id, season, week, depth_position) %>% 
  fill(names(.), .direction = "downup") %>% 
  slice_head(n = 1) %>% 
  ungroup()
```

Let's check again for duplicate `id` instances:

```{r}
nfl_depthCharts %>% 
  group_by(gsis_id, season, week, depth_position) %>% 
  filter(n() > 1)
```

@@

The `nfl_pbp` object is in `game_id`-`drive`-`play_id` form.
That is, each row should be uniquely identified by the combination of `game_id`, `drive`, `play_id`.
Let's rearrange the data accordingly:

```{r}
nfl_pbp <- nfl_pbp %>% 
  select(game_id, drive, play_id, everything()) %>% 
  arrange(game_id, drive, play_id)
```

Let's check for duplicate `id` instances:

```{r}
nfl_pbp %>% 
  group_by(game_id, drive, play_id) %>% 
  filter(n() > 1)
```

@@

player_stats_weekly_kicking <- player_stats_weekly_kicking %>% 
  dplyr::left_join(
    years_of_experience_offense[,c("player_id","season","years_of_experience")],
    by = c("player_id","season")
  )

# Arrange data
player_stats_weekly_offense <- player_stats_weekly_offense %>% 
  arrange(player_display_name, season, week)

player_stats_weekly_defense <- player_stats_weekly_defense %>% 
  arrange(player_display_name, season, week)

player_stats_weekly_kicking <- player_stats_weekly_kicking %>% 
  arrange(player_display_name, season, week)
```

@@

player_stats_seasonal_kicking <- player_stats_seasonal_kicking %>% 
  dplyr::left_join(
    years_of_experience_offense[,c("player_id","season","years_of_experience")],
    by = c("player_id","season")
  )

# Arrange data
player_stats_seasonal_offense <- player_stats_seasonal_offense %>% 
  relocate(season, .after = player_id) %>% 
  arrange(player_display_name, season)

player_stats_seasonal_defense <- player_stats_seasonal_defense %>% 
  relocate(season, .after = player_id) %>% 
  arrange(player_display_name, season)

player_stats_seasonal_kicking <- player_stats_seasonal_kicking %>% 
  relocate(season, .after = player_id) %>% 
  arrange(player_display_name, season)
```

### 2024-07-31 machine-learning.qmd
20240731 - duplicate instances

```{r}
library("LongituRF")
library("doParallel")
library("missForest")
library("powerjoin")
library("tidyverse")
```

@@

```{r}
# Prepare data for merging
## Use common name for the same (gsis_id) ID variable
nfl_actualStats_offense_weekly <- nfl_actualStats_offense_weekly %>% 
  rename(gsis_id = player_id)

@@

## Rename variables
#-todo: rename variables in expected fantasy points so they don't get coalesced with actual points
nfl_draftPicks <- nfl_draftPicks %>%
  rename(
    games_career = games,
    pass_completions_career = pass_completions,
    pass_attempts_career = pass_attempts,
    pass_yards_career = pass_yards,
    pass_tds_career = pass_tds,
    pass_ints_career = pass_ints,
    rush_atts_career = rush_atts,
    rush_yards_career = rush_yards,
    rush_tds_career = rush_tds,
    receptions_career = receptions,
    rec_yards_career = rec_yards,
    rec_tds_career = rec_tds,
    def_solo_tackles_career = def_solo_tackles,
    def_ints_career = def_ints,
    def_sacks_career = def_sacks
  )

# Create lists of objects to merge, depending on data structure: id; or id-season; or id-season-week
#-todo: remove redundant variables
idListToMerge <- list(
  nfl_players %>% filter(!is.na(gsis_id)) %>% select(-season),
  nfl_draftPicks %>% filter(!is.na(gsis_id)) %>% select(-season, -age)
)

@@

seasonalAndWeeklyData <- powerjoin::power_full_join(
  idSeasonWeekMerged,
  seasonalData,
  by = c("gsis_id","season"),
  conflict = coalesce_xy
)
```

@@

seasonalData_subset <- seasonalData %>% 
  dplyr::select(-c(
    birth_date, loaded, full_name, player_name, player_display_name, display_name, suffix, headshot_url, player, pos,
    gsis_id, espn_id, sportradar_id, yahoo_id, rotowire_id, pff_id, fantasy_data_id, sleeper_id, pfr_id,
    pfr_player_id, cfb_player_id, pfr_player_name, esb_id, gsis_it_id, smart_id,
    college, college_name, team_abbr, current_team_id, college_conference, draft_club, status_description_abbr,
    status_short_description, short_name, headshot, uniform_number, jersey_number, first_name, last_name,
    football_name, team
    ))
```

@@

@@

seasonalData_subset_imp <- missForest::missForest(
  seasonalData_subset %>% data.frame,
  verbose = TRUE,
  mtry = 100,
  parallelize = "variables")
```

>>>>

```{r}
library("LongituRF")
library("doParallel")
library("missForest")
library("powerjoin")
library("mice")
library("miceadds")
library("tidyverse")
```

@@

### Prepare Data for Merging {#sec-machineLearningPrepareDataForMerging}

```{r}
# Prepare data for merging
#-todo: calculate years_of_experience
## Use common name for the same (gsis_id) ID variable
nfl_actualStats_offense_weekly <- nfl_actualStats_offense_weekly %>% 
  rename(gsis_id = player_id)

@@

## Rename variables
#-todo: rename variables in expected fantasy points so they don't get coalesced with actual points
nfl_draftPicks <- nfl_draftPicks %>%
  rename(
    games_career = games,
    pass_completions_career = pass_completions,
    pass_attempts_career = pass_attempts,
    pass_yards_career = pass_yards,
    pass_tds_career = pass_tds,
    pass_ints_career = pass_ints,
    rush_atts_career = rush_atts,
    rush_yards_career = rush_yards,
    rush_tds_career = rush_tds,
    receptions_career = receptions,
    rec_yards_career = rec_yards,
    rec_tds_career = rec_tds,
    def_solo_tackles_career = def_solo_tackles,
    def_ints_career = def_ints,
    def_sacks_career = def_sacks
  )
```

```{r}
# Create lists of objects to merge, depending on data structure: id; or id-season; or id-season-week
#-todo: remove redundant variables
idListToMerge <- list(
  nfl_players %>% filter(!is.na(gsis_id)) %>% select(-season, -years_of_experience),
  nfl_draftPicks %>% filter(!is.na(gsis_id)) %>% select(-season, -age)
)

@@

seasonalAndWeeklyData <- powerjoin::power_full_join(
  idSeasonWeekMerged,
  seasonalData,
  by = c("gsis_id","season"),
  conflict = coalesce_xy
)

# check on duplicate id-by-season data (nfl_rosters, nfl_advancedStatsPFR_seasonal)
seasonalData %>% 
  select(gsis_id, season)

seasonalData %>% 
  select(gsis_id, season) %>% 
  unique()
```

@@

seasonalData_subset <- seasonalData %>% 
  dplyr::select(-c(
    birth_date, loaded, full_name, player_name, player_display_name, display_name, suffix, headshot_url, player, pos,
    espn_id, sportradar_id, yahoo_id, rotowire_id, pff_id, fantasy_data_id, sleeper_id, pfr_id,
    pfr_player_id, cfb_player_id, pfr_player_name, esb_id, gsis_it_id, smart_id,
    college, college_name, team_abbr, current_team_id, college_conference, draft_club, status_description_abbr,
    status_short_description, short_name, headshot, uniform_number, jersey_number, first_name, last_name,
    football_name, team
    ))
```

@@

(CONSIDER WIDENING THE DATA TO AVOID MULTILEVEL IMPUTATION)

#### `missForest`

@@

seasonalData_subset_imp <- missForest::missForest(
  seasonalData_subset %>% select(ageCentered20, ageCentered20Quadratic, fantasy_points) %>% data.frame,
  verbose = TRUE,
  mtry = 100,
  parallelize = "forests")
```

#### `MICE`

```{r}
# Specify data to impute
dataToImpute <- seasonalData_subset %>% 
  select(gsis_id, season, ageCentered20, ageCentered20Quadratic, fantasy_points, years_of_experience) %>% 
  filter(!is.na(gsis_id), !is.na(season)) %>% 
  mutate(ID = as.integer(as.factor(as.character(gsis_id)))) %>% 
  arrange(gsis_id, season)

# Specify variables to impute
varsToImpute <- c("fantasy_points")
Y <- varsToImpute

# Specify predictors of missing data
clusterVariable <- "ID"
fixedEffectPredictors <- c("ageCentered20Quadratic")
randomEffectPrectors <- c("ageCentered20")

# Specify number of imputations
numImputations <- 1

# Specify the imputation method
meth <- make.method(dataToImpute)
meth[1:length(meth)] <- ""
meth[Y] <- "2l.pmm" # specify the imputation method here; this can differ by outcome variable

# Specify the predictor matrix

pred <- make.predictorMatrix(dataToImpute)
pred[1:nrow(pred), 1:ncol(pred)] <- 0
pred[Y, clusterVariable] <- (-2) # cluster variable
pred[Y, randomEffectPrectors] <- 2 # random effect predictor
#pred[Y, fixedEffectPredictors] <- 1 # fixed effect predictor
pred[Y, Y] <- 1 # fixed effect predictor of each predictor in predicting the other predictors

diag(pred) <- 0
pred

# Imputation
mi_mice <- mice(
  as.data.frame(dataToImpute),
  method = meth,
  predictorMatrix = pred,
  m = numImputations,
  maxit = 100, # generally use 100 maximum iterations; this example uses 5 for speed
  seed = 52242)

mi_mice$loggedEvents

seasonalData_subset_imp <- complete(
  mi_mice,
  action = "stacked",
  include = FALSE)
```
